Making Sequential Programs Deterministic

### 2019-03-28 design.tex
Fix LaTeX error, complete figure
Edits in section 2.2

    $\infer[\textsc{invariant type param}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

    \caption{A sampling of standard type checking rules.  Most standard rules are omitted
    for brevity.  $\sqsubseteq$ is overloaded for classes and types.}
    \label{typecheck-rules}
\end{figure}

@@

    \caption{Type checking rules.  $\sqsubseteq$ is overloaded for classes,
    basetypes, type qualifiers, and types.  Most standard rules are omitted
    for brevity.}
    \label{typecheck-rules}
\end{figure}

@@

A map is a dictionary or associative array, such as a hash table.
Our type system largely treats a map as a collection of key--value pairs.  Like any
collection, a map may be \<Det>, \<OrderNonDet>, or \<NonDet>.

@@

\subsection{Additional typing rules for collection elements}\label{collection-rules}

@@

Figure~\ref{type-validity} shows the type validity\todo{The caption says
  ``type well formedness''.  Choose one term and use it consistently
  throughout.}
rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
where $\kappa$ is a type qualifier from the determinism type hierarchy and $\beta$ is a Java type.
As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet> can be written on any Java type.
The (determinism) type of a collection must be a supertype or equal to
its element type (Rule $\textsc{ordernondet collection}$) (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<NonDet>, then the type parameter may not be
\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in Figure~\ref{type-validity}). Although such types could exist, they are
disallowed to prevent the following bug:

@@

%\todo{Fewer examples, less variation.  Use \<List> and \<String>.}

Some examples of valid types are:\todo{double-check}
\begin{itemize}
    \item \codeid{NonDet\ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{OrderNonDet List<OrderNonDet Set<...>\relax >}
    \item \codeid{OrderNonDet List<Det Integer>}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
\end{itemize}

These types are invalid:
\begin{itemize}
    \item \codeid{NonDet\ \ \ \ \ \ List<OrderNonDet Integer>}
    \item \codeid{NonDet\ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
    \item \codeid{OrderNonDet List<NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Set<...>\relax >}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<OrderNonDet Set<...>\relax >}
\end{itemize}

>>>>

    $\infer[\textsc{invariant type param}]{C_1\angles{\overline{\tau_i}}
      \sqsubseteq C_2\angles{\overline{\tau_i}}}{C_1 \sqsubseteq C_2}$

    \caption{A sampling of standard type checking rules.  Most standard rules are omitted
    for brevity.  $\sqsubseteq$ is overloaded for classes and types.}
    \label{typecheck-rules-standard}
\end{figure}

@@

    \caption{Type checking rules.  $\sqsubseteq$ is overloaded for classes,
    basetypes, type qualifiers, and types.}
    \label{typecheck-rules}
\end{figure}

@@

A map is a dictionary or associative array, such as a hash table.
Our type system largely treats a map as a collection of key--value pairs.  Like a
collection, a map may be \<Det>, \<OrderNonDet>, or \<NonDet>.  (In Java,
\<HashMap> cannot be deterministic, but \<TreeMap> and \<LinkedHashMap>
can.\todo{forward reference?})

@@

\subsection{Collection types}\label{collection-rules}

@@

Figure~\ref{type-validity} shows the type validity\todo{The caption says
  ``type well formedness''.  Choose one term and use it consistently
  throughout.}
rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
\<Det>) and $\beta$ is a Java basetype.
As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet>
can be written on any Java basetype.\todo{As noted in figure caption, that's
  not true, in some contexts such as a type argument.}

The determinism type qualifier on a collection must be a supertype or equal to
its element type qualifier (Rule $\textsc{ordernondet collection}$ in Figure~\ref{fig-determinism-collections}).
If the Collection is \<NonDet>, then the type argument may not be
\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in Figure~\ref{type-validity}). Although such types could exist, they are
disallowed to prevent the following bug:

@@

\todo{Some notes:}
Typical type for \<get> is \\
$\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau$ \\
but we need to change that.  If either argument is \<OrderNonDet> or \<NonDet>, then the
result is \<NonDet>.
If both arguments are \<Det>, the result is $\tau$ as usual.
\todo{In that case, we know that $\tau = \<Det> \beta$.}

\smallskip
\noindent
\begin{minipage}{.48\textwidth}
Some examples of valid types are:
% Using "~" instead of "\ " does not work within \codeid :-(
\begin{smaller}
\begin{Verbatim}
@NonDet      List<@NonDet      Integer>
@OrderNonDet List<@OrderNonDet Set<...>>
@OrderNonDet List<@Det         Integer>
@Det         List<@Det         Integer>

\end{Verbatim}
\end{smaller}
\end{minipage}
\hfill
\begin{minipage}{.46\textwidth}
These types are invalid:
\begin{smaller}
\begin{Verbatim}
@NonDet      List<@OrderNonDet Set<...>>
@NonDet      List<@Det         Integer>
@OrderNonDet List<@NonDet      Integer>
@Det         List<@NonDet      Integer>
@Det         List<@OrderNonDet Set<...>>
\end{Verbatim}
\end{smaller}
\end{minipage}
\smallskip
